---
pagetitle: "Meet the WNBA"
description: "A season of WNBA basketball, one player at a time."
format:
  html:
    theme: courtside.scss
    page-layout: custom
    toc: false
    anchor-sections: false
    embed-resources: true
    link-external-newwindow: true
knitr:
  opts_chunk:
    echo: false
    warning: false
    message: false
---

```{r packages}
library(dplyr)
library(tidyr)
library(purrr)
library(stringr)
library(lubridate)
library(jsonlite)
library(wehoop)
# required by wehoop: pre-2024 season .rds files store their columns as arrow
# ALTREP vectors, which readRDS silently returns as length-zero vectors (and
# rbindlist then pads to NA) unless arrow is loaded
library(arrow)
```

```{r data}
this_year <- as.integer(year(now()))
seasons <- 2004:this_year

config <- config::get()

if (config$data_source == "local") {
  dat <- readRDS("dat.rds")
  sched <- readRDS("sched.rds")
}

if (config$data_source == "wehoop") {
  dat <- load_wnba_player_box(seasons = seasons)
  sched <- load_wnba_schedule(seasons = seasons)
}

game_types <- sched |> select(game_id, type_id)

# The two filters below are the load-bearing ones, and METHODOLOGY.md explains why.
# type_id != 4 drops All-Star exhibitions, whose rosters are not real teams.
# The did_not_play / is.na(points) pair keeps "one row = one game actually
# played" true, which every per-game average on this page depends on.
dat <- dat |>
  left_join(game_types, by = "game_id") |>
  filter(type_id != 4) |>
  filter(!(did_not_play %in% TRUE), !is.na(points))

# regular season only, current year — the whole page except the rookie flag
now_dat <- dat |>
  filter(season == this_year, type_id == 1, !is.na(athlete_display_name))

MIN_GAMES <- 10L
```

```{r helpers}
# ESPN's brand colours are chosen for white backgrounds, so several of them
# (Dallas navy, Indiana navy, Seattle green) disappear on this page. Lift value
# and saturation to a floor so every franchise reads as itself against the ink.
accent <- function(hex) {
  hex <- ifelse(is.na(hex) | hex == "", "f57b20", hex)
  map_chr(hex, function(h) {
    rgb <- col2rgb(paste0("#", h))
    hsv <- rgb2hsv(rgb)
    v <- max(hsv[3, 1], 0.62)
    s <- if (hsv[2, 1] < 0.06) 0 else max(hsv[2, 1], 0.42)
    hsv(hsv[1, 1], s, v)
  })
}

fmt_date <- function(d) {
  d <- as.Date(d)
  paste0(format(d, "%b"), " ", as.integer(format(d, "%d")), ", ", format(d, "%Y"))
}

short_date <- function(d) {
  d <- as.Date(d)
  paste0(format(d, "%b"), " ", as.integer(format(d, "%d")))
}

# rank within the qualified pool only, so a one-game cameo cannot lead the
# league in points per game
rank_of <- function(x, keep) {
  r <- rep(NA_integer_, length(x))
  r[keep] <- rank(-x[keep], ties.method = "min")
  r
}

pct <- function(made, att) if_else(att > 0, 100 * made / att, NA_real_)

# 1st, 2nd, 3rd, 4th … 11th, 12th, 13th
ord_en <- function(n) {
  suffix <- c("th", "st", "nd", "rd", rep("th", 6))
  paste0(n, if_else(n %% 100 %in% 11:13, "th", suffix[(n %% 10) + 1]))
}

# ESPN serves headshots at 350px and crests at 500px — 310 KB and 35 KB each,
# for pictures this page draws at 132px and 30px. Its own combiner endpoint
# resizes on the CDN, which takes the wall from roughly 70 MB of portraits to
# 9 MB and the crests, which pandoc inlines into the file, from 1 MB to 90 KB.
espn_img <- function(url, w, h = w) {
  path <- str_replace(url, "^https?://[a-z0-9.]+", "")
  if_else(
    is.na(url), NA_character_,
    paste0("https://a.espncdn.com/combiner/i?img=", path, "&w=", w, "&h=", h)
  )
}

# most common non-missing value, or NA when a player has none recorded
mode_chr <- function(x) {
  x <- x[!is.na(x)]
  if (!length(x)) return(NA_character_)
  names(sort(table(x), decreasing = TRUE))[1]
}
```

```{r players}
rookie_year <- dat |>
  summarise(first_season = min(season, na.rm = TRUE), .by = athlete_display_name)

players <- now_dat |>
  summarise(
    shot = first(na.omit(athlete_headshot_href)),
    jersey = first(na.omit(athlete_jersey)),
    pos = mode_chr(athlete_position_abbreviation),
    team = team_abbreviation[which.max(game_date)],
    teamName = team_display_name[which.max(game_date)],
    color = team_color[which.max(game_date)],
    logo = team_logo[which.max(game_date)],
    g = n_distinct(game_id),
    min = sum(minutes, na.rm = TRUE),
    pts = sum(points, na.rm = TRUE),
    reb = sum(rebounds, na.rm = TRUE),
    ast = sum(assists, na.rm = TRUE),
    stl = sum(steals, na.rm = TRUE),
    blk = sum(blocks, na.rm = TRUE),
    tov = sum(turnovers, na.rm = TRUE),
    fgm = sum(field_goals_made, na.rm = TRUE),
    fga = sum(field_goals_attempted, na.rm = TRUE),
    tpm = sum(three_point_field_goals_made, na.rm = TRUE),
    tpa = sum(three_point_field_goals_attempted, na.rm = TRUE),
    ftm = sum(free_throws_made, na.rm = TRUE),
    fta = sum(free_throws_attempted, na.rm = TRUE),
    # a double-double is two of the five counting stats in double figures on
    # the same night. A triple-double is three of them
    dd = sum(rowSums(cbind(points, rebounds, assists, steals, blocks) >= 10, na.rm = TRUE) >= 2),
    td = sum(rowSums(cbind(points, rebounds, assists, steals, blocks) >= 10, na.rm = TRUE) >= 3),
    .by = athlete_display_name
  ) |>
  rename(name = athlete_display_name) |>
  left_join(rookie_year, by = join_by(name == athlete_display_name)) |>
  mutate(
    first = word(name, 1),
    last = word(name, -1),
    initials = paste0(str_sub(first, 1, 1), str_sub(last, 1, 1)),
    rook = first_season == this_year,
    shot = espn_img(shot, 220, 160),
    logo = espn_img(logo, 64),
    color = accent(color),
    mpg = min / g, ppg = pts / g, rpg = reb / g, apg = ast / g,
    spg = stl / g, bpg = blk / g, tpg = tpm / g,
    fgp = pct(fgm, fga), tpp = pct(tpm, tpa), ftp = pct(ftm, fta),
    ts = pct(pts, 2 * (fga + 0.44 * fta))
  )

qual <- players$g >= MIN_GAMES

players <- players |>
  mutate(
    rk_ppg = rank_of(ppg, qual), rk_rpg = rank_of(rpg, qual),
    rk_apg = rank_of(apg, qual), rk_spg = rank_of(spg, qual),
    rk_bpg = rank_of(bpg, qual)
  ) |>
  arrange(desc(pts))

# one bar per game in the dossier's game log
game_log <- now_dat |>
  arrange(athlete_display_name, game_date) |>
  transmute(
    name = athlete_display_name,
    d = short_date(game_date),
    o = opponent_team_abbreviation,
    p = as.integer(replace_na(points, 0)),
    w = as.integer(team_winner %in% TRUE)
  ) |>
  nest(gl = c(d, o, p, w))

players <- players |>
  left_join(game_log, by = "name") |>
  select(
    name, first, last, initials, jersey, pos, team, teamName, color, logo, shot, rook,
    g, min, mpg, pts, ppg, reb, rpg, ast, apg, stl, spg, blk, bpg, tov, tpm, tpg,
    fgm, fga, fgp, tpa, tpp, ftm, fta, ftp, ts, dd, td,
    starts_with("rk_"), gl
  ) |>
  mutate(across(c(mpg, ppg, rpg, apg, spg, bpg, tpg, fgp, tpp, ftp, ts), \(x) round(x, 1)))
```

```{r teams}
team_games <- now_dat |>
  distinct(game_id, game_date, team_abbreviation, team_display_name,
           team_color, team_logo, team_winner, team_score, opponent_team_score)

teams <- team_games |>
  arrange(game_date) |>
  summarise(
    name = last(team_display_name),
    color = accent(last(team_color)),
    logo = espn_img(last(team_logo), 64),
    w = sum(team_winner %in% TRUE),
    l = sum(team_winner %in% FALSE),
    pf = mean(team_score, na.rm = TRUE),
    pa = mean(opponent_team_score, na.rm = TRUE),
    form = list(tail(if_else(team_winner %in% TRUE, "w", "l"), 10)),
    .by = team_abbreviation
  ) |>
  rename(abbr = team_abbreviation) |>
  mutate(pct = w / (w + l)) |>
  arrange(desc(pct), desc(w))

# most recent finals, for the hero ticker
recent <- team_games |>
  left_join(
    now_dat |> distinct(game_id, team_abbreviation, home_away),
    by = c("game_id", "team_abbreviation")
  ) |>
  select(game_id, game_date, abbr = team_abbreviation, score = team_score,
         color = team_color, home_away, team_winner) |>
  arrange(desc(game_date), game_id, home_away) |>
  mutate(side = tolower(home_away)) |>
  distinct(game_id, side, .keep_all = TRUE) |>
  pivot_wider(
    id_cols = c(game_id, game_date),
    names_from = side,
    values_from = c(abbr, score, color, team_winner)
  ) |>
  arrange(desc(game_date)) |>
  head(14)
```

```{r race}
# One line per player: points scored so far, at every date the league played.
# Values are carried forward between a player's own games so the lines run
# level rather than dropping to zero on their nights off, and are null before
# a player's debut so nobody appears to start the season already flat.
dates <- sort(unique(as_date(now_dat$game_date)))

# `gd`, not `d`: a column named `d` would partial-match complete()'s own `data`
# argument, since it sits ahead of the dots.
race_long <- now_dat |>
  mutate(gd = as_date(game_date)) |>
  summarise(p = sum(points, na.rm = TRUE), .by = c(athlete_display_name, gd)) |>
  complete(athlete_display_name, gd = dates, fill = list(p = 0)) |>
  arrange(athlete_display_name, gd) |>
  mutate(cum = cumsum(p), .by = athlete_display_name)

debut <- now_dat |>
  summarise(debut = min(as_date(game_date)), .by = athlete_display_name)

race_long <- race_long |>
  left_join(debut, by = "athlete_display_name") |>
  mutate(cum = if_else(gd < debut, NA_integer_, as.integer(cum)))

top8 <- players$name[1:8]

race_series <- race_long |>
  arrange(athlete_display_name, gd) |>
  summarise(v = list(cum), .by = athlete_display_name) |>
  rename(n = athlete_display_name) |>
  left_join(players |> select(n = name, c = color, pts), by = "n") |>
  mutate(top = n %in% top8) |>
  arrange(pts) |> # leaders drawn last, on top of the pack
  select(n, c, top, v)

# roughly monthly x-axis ticks
tick_i <- which(!duplicated(format(dates, "%Y-%m")))
race_ticks <- tibble(i = tick_i - 1L, l = toupper(format(dates[tick_i], "%b")))
```

```{r career}
# Career totals back to 2004, regular season only. The page is otherwise about
# this season alone, and a newcomer has no way to tell whether 746 points is a
# lot without something to measure it against.
data_from <- min(dat$season)

career <- dat |>
  filter(type_id == 1, !is.na(athlete_display_name)) |>
  arrange(game_date) |>
  summarise(
    shot = last(na.omit(athlete_headshot_href)),
    team = last(team_abbreviation),
    color = accent(last(team_color)),
    first_season = min(season), last_season = max(season),
    n_seasons = n_distinct(season),
    g = n_distinct(game_id),
    pts = sum(points, na.rm = TRUE),
    reb = sum(rebounds, na.rm = TRUE),
    ast = sum(assists, na.rm = TRUE),
    stl = sum(steals, na.rm = TRUE),
    blk = sum(blocks, na.rm = TRUE),
    tpm = sum(three_point_field_goals_made, na.rm = TRUE),
    .by = athlete_display_name
  ) |>
  rename(name = athlete_display_name) |>
  mutate(
    shot = espn_img(shot, 220, 160),
    active = last_season == this_year,
    initials = paste0(str_sub(word(name, 1), 1, 1), str_sub(word(name, -1), 1, 1)),
    span = if_else(
      first_season == last_season,
      as.character(first_season),
      paste0(first_season, "&ndash;", last_season)
    ),
    # The league began in 1997 but this data begins in 2004, so anyone whose
    # first season here is 2004 may well have been playing before it. Their
    # span and their totals are both cut off, and saying "2004-2016" of a
    # career that ran 2002-2016 would be a claim the data cannot support.
    trunc = first_season == data_from
  )

career_metrics <- c(pts = "Points", reb = "Rebounds", ast = "Assists",
                    stl = "Steals", blk = "Blocks", tpm = "3-Pointers", g = "Games")

career_boards <- map(
  set_names(names(career_metrics)),
  \(k) career |>
    slice_max(.data[[k]], n = 10, with_ties = FALSE) |>
    transmute(name, team, color, shot, initials, span, trunc, active, v = .data[[k]])
)

# does any leaderboard on show actually contain a truncated career?
career_has_trunc <- any(map_lgl(career_boards, \(b) any(b$trunc)))
```

```{r payload}
#| output: asis

league_avg <- now_dat |>
  summarise(
    fgp = pct(sum(field_goals_made, na.rm = TRUE), sum(field_goals_attempted, na.rm = TRUE)),
    tpp = pct(sum(three_point_field_goals_made, na.rm = TRUE), sum(three_point_field_goals_attempted, na.rm = TRUE)),
    ftp = pct(sum(free_throws_made, na.rm = TRUE), sum(free_throws_attempted, na.rm = TRUE)),
    # two-pointers only: field goals are the two and the three together, so the
    # decoder cannot compare a three-point percentage against `fgp` and claim
    # to be comparing it against shots from inside the arc
    twop = pct(
      sum(field_goals_made, na.rm = TRUE) - sum(three_point_field_goals_made, na.rm = TRUE),
      sum(field_goals_attempted, na.rm = TRUE) - sum(three_point_field_goals_attempted, na.rm = TRUE)
    )
  ) |>
  mutate(across(everything(), \(x) round(x, 1)))

# points each shot type returns per attempt, which is the whole reason a lower
# three-point percentage is not the same as worse shooting
pts_per_three <- sprintf("%.2f", 3 * league_avg$tpp / 100)
pts_per_two <- sprintf("%.2f", 2 * league_avg$twop / 100)

league_td <- sum(players$td)

payload <- list(
  season = this_year,
  minGames = MIN_GAMES,
  qualified = sum(qual),
  opening = players$name[1],
  avg = as.list(league_avg),
  players = players,
  career = career_boards,
  race = list(
    dates = format(dates, "%Y-%m-%d"),
    ticks = race_ticks,
    max = max(race_long$cum, na.rm = TRUE),
    series = race_series
  )
)

json <- toJSON(payload, dataframe = "rows", auto_unbox = TRUE, na = "null", digits = 4)
# a literal </script> inside a JSON string would end the tag early. Nothing in
# this data contains one today, but the escape costs nothing
json <- gsub("</", "<\\\\/", json, fixed = TRUE)

cat('<script id="payload" type="application/json">', json, "</script>\n", sep = "")
```

```{r furniture}
#| output: asis

n_games <- n_distinct(now_dat$game_id)
n_points <- sum(now_dat$points, na.rm = TRUE)
data_through <- fmt_date(max(now_dat$game_date, na.rm = TRUE))

h <- function(...) cat(..., "\n", sep = "")

h('<div id="progress"></div><div id="tip"></div>')

# the same names the headings carry, so the rail never gives a section a
# second name
rail <- c(tipoff = "Tip-off", who = "Who's who", leaders = "The pecking order",
          `race-band` = "The race", teams = "Fifteen teams",
          record = "The record book", decoder = "Decoder")
h('<nav id="rail">', paste0(
  '<a href="#', names(rail), '"><span class="lbl">', rail, '</span><span class="dot"></span></a>',
  collapse = ""
), "</nav>")
```

::: {#tipoff .band}

```{r court}
#| output: asis

h(
  '<svg id="court" viewBox="0 0 940 500" preserveAspectRatio="xMidYMid meet" aria-hidden="true">',
  '<rect class="ln" x="10" y="10" width="920" height="480"/>',
  '<line class="ln" x1="470" y1="10" x2="470" y2="490"/>',
  '<circle class="ln" cx="470" cy="250" r="60"/>',
  '<rect class="ln" x="10" y="170" width="190" height="160"/>',
  '<rect class="ln" x="740" y="170" width="190" height="160"/>',
  '<circle class="ln" cx="200" cy="250" r="60"/>',
  '<circle class="ln" cx="740" cy="250" r="60"/>',
  '<circle class="ln" cx="62" cy="250" r="9"/>',
  '<circle class="ln" cx="868" cy="250" r="9"/>',
  '<path class="ln" d="M 10 30 L 90 30 A 221 221 0 0 1 90 470 L 10 470"/>',
  '<path class="ln" d="M 930 30 L 850 30 A 221 221 0 0 0 850 470 L 930 470"/>',
  "</svg>"
)
```

::: {.hero}

```{r hero-head}
#| output: asis

h('<div class="hero-kicker"><span class="live">', this_year, ' Regular Season</span>',
  '<span>Data through ', data_through, "</span></div>")
h('<h1 class="hero-title">Meet<span class="l2">the <span class="ghost">WNBA</span></span></h1>')
```

::: {.hero-lede}
New to the league? Start anywhere. **One player is selected at a time, and
picking her anywhere picks her everywhere** — from the wall of cards, a
leaderboard, the record book or the race chart. Her panel opens beside the
wall. The decoder at the foot of the page explains what each number means.
:::

```{r hero-figures}
#| output: asis

figure <- function(v, l) paste0(
  '<div class="figure"><span class="fv" data-to="', v, '">0</span><span class="fl">', l, "</span></div>"
)

h('<div class="hero-figures">',
  figure(nrow(players), "Players"), figure(nrow(teams), "Teams"),
  figure(n_games, "Games played"), figure(n_points, "Points scored"),
  "</div>")
```

:::

```{r ticker}
#| output: asis

ticker_item <- function(r) {
  side <- function(ab, sc, col, won) paste0(
    '<span class="side', if (won) " won" else "", '" style="--tc:', accent(col), '">',
    '<i class="swatch"></i><span class="ab">', ab, '</span><span class="sc">', sc, "</span></span>"
  )
  paste0(
    '<div class="final">',
    side(r$abbr_away, r$score_away, r$color_away, isTRUE(r$team_winner_away)),
    '<span class="at">at</span>',
    side(r$abbr_home, r$score_home, r$color_home, isTRUE(r$team_winner_home)),
    '<span class="day">', toupper(short_date(r$game_date)), "</span></div>"
  )
}

finals <- paste0(map_chr(seq_len(nrow(recent)), \(i) ticker_item(recent[i, ])), collapse = "")

h('<div class="ticker"><div class="ticker-track">', finals, finals, "</div></div>")
```

:::

::: {#who .band}

::: {.band-head .reveal}
## Who's who {.band-title}

::: {.band-sub}
All **`r nrow(players)` players** who have played at least one game this season.
Sort them by scoring, rebounds, assists, minutes or alphabetically, and filter
by team. **Click a card and the panel beside the wall fills with that player's
season.** That same player is drawn as the thick pale line in the race chart
further down.
:::
:::

```{r wall}
#| output: asis

chips <- paste0(
  '<button class="chip all" data-team="">All ', nrow(teams), ' teams</button>',
  '<button class="chip" data-rookies="1">Rookies</button>',
  paste0(
    '<button class="chip" data-team="', teams$abbr, '">',
    '<img src="', teams$logo, '" alt="" loading="lazy">', teams$abbr, "</button>",
    collapse = ""
  )
)

sortbar <- paste0(
  '<div class="sortbar" id="wall-sort">',
  paste0(
    '<button data-sort="', c("pts", "reb", "ast", "min", "az"),
    '"', c(" class=\"on\"", "", "", "", ""), ">",
    c("Scoring", "Rebounds", "Assists", "Minutes", "A&ndash;Z"), "</button>",
    collapse = ""
  ),
  "</div>"
)

h('<div class="controls reveal">',
  '<input id="search" type="search" placeholder="Search a name or team" autocomplete="off">',
  sortbar,
  '<span class="band-sub" id="wall-count" style="margin:0;font-size:.8rem"></span>',
  "</div>")
h('<div class="teamchips reveal">', chips, "</div>")
# a <div>, not an <aside>: Quarto's own stylesheet places every <aside> in the
# margin column of its page grid, which pulls it out of this two-column shell
h('<div class="wall-shell">',
  '<div class="wall-col"><div id="wall"></div>',
  '<button class="more" id="more" type="button" hidden></button></div>',
  '<div id="dossier"></div></div>')
```

:::

::: {#leaders .band}

::: {.band-head .reveal}
## The pecking order {.band-title}

::: {.band-sub}
The top ten in each of seven categories. Every bar starts at zero and the
longest one is the leader, so a bar half as long is half the total. **Click a
row to open that player's panel.**
:::
:::

```{r board}
#| output: asis

lead_tabs <- paste0(
  '<div class="tabs" id="lead-tabs">',
  paste0(
    '<button data-m="', c("pts", "reb", "ast", "stl", "blk", "tpm", "dd"),
    '"', c(" class=\"on\"", "", "", "", "", "", ""), ">",
    c("Points", "Rebounds", "Assists", "Steals", "Blocks", "3-Pointers", "Double-Doubles"),
    "</button>",
    collapse = ""
  ),
  "</div>"
)

per_toggle <- paste0(
  '<div class="sortbar" id="per-toggle">',
  '<button data-per="0" class="on">Season total</button>',
  '<button data-per="1">Per game</button>',
  "</div>"
)

h('<div class="lead-controls reveal">', lead_tabs, per_toggle, "</div>")
h('<div id="board" class="reveal"></div>')
h('<p class="band-sub footnote" id="board-note"></p>')
```

:::

::: {#race-band .band}

::: {.band-head .reveal}
## The race {.band-title}

::: {.band-sub}
Every line is one player's points, adding up game by game. The eight highest
scorers are labelled and drawn in their team's colour, and the other
`r nrow(players) - 8` stay grey. Where a line gets steeper, that player was
scoring more per game than she had been. **Click one of the labelled lines to
open that player's panel, and whoever is open there is the thick pale line.**
:::
:::

```{r race-chart}
#| output: asis

h('<div class="race-wrap reveal" id="race-anchor">',
  '<svg id="race" viewBox="0 0 1000 430" role="img" aria-label="Cumulative points by date for every player"></svg>',
  "</div>")
```

:::

::: {#teams .band}

::: {.band-head .reveal}
## Fifteen teams {.band-title}

::: {.band-sub}
Each card carries the won--lost record, the average points scored and allowed
per game, and the last ten results as small squares: green for a win, red for a
loss, oldest on the left. **Click a card to filter the wall of players down to
that team's roster.**
:::
:::

```{r standings}
#| output: asis

tcards <- paste0(
  '<button class="tcard" type="button" style="--team:', teams$color, '" data-team="', teams$abbr, '">',
  '<span class="top"><img src="', teams$logo, '" alt="" loading="lazy">',
  '<span class="nm">', teams$name, "</span>",
  '<span class="rec">', teams$w, "&ndash;", teams$l, "</span></span>",
  '<span class="line2"><span>', sprintf("%.0f", teams$pf), " for &middot; ",
  sprintf("%.0f", teams$pa), " against</span>",
  '<span class="form">',
  map_chr(teams$form, \(f) paste0('<i class="', f, '"></i>', collapse = "")),
  "</span></span></button>",
  collapse = ""
)

h('<div id="standings" class="reveal">', tcards, "</div>")
```

:::

::: {#record .band}

::: {.band-head .reveal}
## The record book {.band-title}

::: {.band-sub}
Everyone who has played a regular-season game since **`r data_from`**, ranked by
career totals. The players in orange are still playing this season, so their
totals keep rising. **Click one of them to open her panel.**
:::
:::

```{r career-board}
#| output: asis

record_tabs <- paste0(
  '<div class="tabs" id="career-tabs">',
  paste0(
    '<button data-m="', names(career_metrics), '"',
    c(" class=\"on\"", rep("", length(career_metrics) - 1)), ">",
    career_metrics, "</button>",
    collapse = ""
  ),
  "</div>"
)

h('<div class="lead-controls reveal">', record_tabs, "</div>")
h('<div id="career-board" class="reveal"></div>')
```

```{r record-footnote}
#| child: !expr if (career_has_trunc) "record-footnote.qmd"
```

:::

```{r decoder-examples}
# Each term is illustrated with a player from this season rather than a made-up
# number, so the vocabulary attaches to someone the reader has just seen.
top_ppg <- players |> filter(!is.na(rk_ppg)) |> slice_min(rk_ppg, n = 1, with_ties = FALSE)
top_tot <- players |> slice_max(pts, n = 1, with_ties = FALSE)
top_dd <- players |> slice_max(dd, n = 1, with_ties = FALSE)
top_ts <- players |> filter(fga >= 150) |> slice_max(ts, n = 1, with_ties = FALSE)
top_tpm <- players |> slice_max(tpm, n = 1, with_ties = FALSE)
top_fill <- players |> filter(g >= MIN_GAMES) |>
  mutate(fill = rpg + apg + spg + bpg) |> slice_max(fill, n = 1, with_ties = FALSE)
top_gp <- players |> slice_max(g, n = 1, with_ties = FALSE)

# The point of the per-game card is that the two orderings disagree, so it
# needs the player they disagree about most — not just the two leaders, who
# this season happen to be the same person. Among the qualified, this is the
# biggest drop from a per-game rank to a season-total rank.
gap <- players |>
  filter(g >= MIN_GAMES) |>
  mutate(rk_tot = rank(-pts, ties.method = "min")) |>
  filter(rk_ppg <= 25) |>
  slice_max(rk_tot - rk_ppg, n = 1, with_ties = FALSE)

# "played every game her team has" is only true if the two counts match
top_gp_team_games <- team_games |>
  filter(team_abbreviation == top_gp$team) |>
  nrow()

# One example per decoder card. The cards themselves are markdown below; only
# these sentences have to be built, because each one quotes this season.
eg <- list(
  per_game = paste0(
    "<b>", top_tot$name, "</b> has scored the most points overall, ",
    top_tot$pts, " in ", top_tot$g, " games",
    if (top_ppg$name != top_tot$name) {
      paste0(", while <b>", top_ppg$name, "</b> leads per game at ",
             sprintf("%.1f", top_ppg$ppg), ". ")
    } else {
      paste0(", and she leads per game too, at ", sprintf("%.1f", top_ppg$ppg),
             ". The two rankings diverge further down. ")
    },
    "<b>", gap$name, "</b> is ", ord_en(gap$rk_ppg), " at ",
    sprintf("%.1f", gap$ppg), " a game but has played only ", gap$g,
    " games, which leaves her ", ord_en(gap$rk_tot), " in total points."
  ),
  double_double = paste0(
    "<b>", top_dd$name, "</b> has ", top_dd$dd, " double-doubles this season. ",
    "The whole league has produced ", league_td, " triple-double",
    if (league_td == 1) "" else "s", " so far."
  ),
  contributions = paste0(
    "<b>", top_fill$name, "</b> has the highest combined total of the four: ",
    sprintf("%.1f", top_fill$rpg), " rebounds, ", sprintf("%.1f", top_fill$apg),
    " assists, ", sprintf("%.1f", top_fill$spg), " steals and ",
    sprintf("%.1f", top_fill$bpg), " blocks a game."
  ),
  shooting = paste0(
    "The league makes ", sprintf("%.1f%%", league_avg$tpp), " of its threes and ",
    sprintf("%.1f%%", league_avg$twop), " of its two-point shots, which come to ",
    pts_per_three, " and ", pts_per_two, " points per attempt. The ",
    sprintf("%.1f", league_avg$twop - league_avg$tpp),
    "-point gap between the percentages is worth almost nothing. <b>", top_tpm$name,
    "</b> has made the most threes, ", top_tpm$tpm, " at ",
    sprintf("%.1f%%", top_tpm$tpp), "."
  ),
  true_shooting = paste0(
    "<b>", top_ts$name, "</b> is the most efficient of the players who have attempted ",
    "150 or more field goals, at ", sprintf("%.1f%%", top_ts$ts), "."
  ),
  games_played = paste0(
    "<b>", top_gp$name, "</b> ",
    if (top_gp$g >= top_gp_team_games) {
      paste0("has played every game her team has this season, all ", top_gp$g)
    } else {
      paste0("has played the most games of anyone, ", top_gp$g)
    },
    ", for ", round(top_gp$min), " minutes on court."
  )
)
```

<!-- the heading below is given an id, because "Decoder" would generate
     #decoder on its own and collide with the band that carries it -->

::: {#decoder .band}

::: {.band-head .reveal}
## Decoder {#decoder-title .band-title}

::: {.band-sub}
Six terms used throughout this page, each one shown with a player from this
season rather than an invented example.
:::
:::

::: {#decoder-grid .reveal}

::: {.term}
### Per game vs total

A season total favours whoever played the most games. A per-game average
favours whoever produced the most in each game. Missed games pull the two
rankings apart, which is why each leaderboard offers both.

::: {.eg}
`r eg$per_game`
:::
:::

::: {.term}
### Double-double

Ten or more in two of the five counting statistics in one game: points,
rebounds, assists, steals and blocks. Most pair points with rebounds. Ten or
more in three of them is a triple-double.

::: {.eg}
`r eg$double_double`
:::
:::

::: {.term}
### Rebounds, assists, steals, blocks

A **rebound** is a missed shot collected off the rim or backboard. An
**assist** is a pass that leads directly to a team-mate's basket. A **steal**
takes the ball away from an opponent, and a **block** deflects an opponent's
shot. These four measure what a player contributes apart from scoring.

::: {.eg}
`r eg$contributions`
:::
:::

::: {.term}
### Shooting percentages

The share of a player's shots that go in. A shot from behind the three-point
arc, roughly 22 feet from the basket, is worth three points rather than two, so
a percentage from outside the arc and one from inside it cannot be compared
directly. Multiply each by what that shot is worth.

::: {.eg}
`r eg$shooting`
:::
:::

::: {.term}
### True shooting %

A single percentage that counts a three as three points and includes free
throws, so it measures scoring efficiency across every kind of shot. It
normally runs higher than field-goal percentage, so compare a player's true
shooting only with another player's true shooting.

::: {.eg}
`r eg$true_shooting`
:::
:::

::: {.term}
### Games played

Every per-game figure divides by this number, so it sets what all the others
mean. A player who missed a month can lead a per-game ranking without having
been the season's best. Per-game rankings on this page require
**`r MIN_GAMES` games**, but season totals have no minimum.

::: {.eg}
`r eg$games_played`
:::
:::

:::

:::

::: {#fineprint .band}

<details>
<summary>The fine print: how these numbers were counted</summary>

::: {.method-body}

{{< include METHODOLOGY.md >}}

:::

</details>

```{r colophon}
#| output: asis
cat(
  '<div class="colophon">',
  '<span>Meet the WNBA</span>',
  '<span>Data through ', data_through, "</span>",
  '<span>Rendered ', fmt_date(Sys.Date()), "</span>",
  '<span>Source: <a href="https://wehoop.sportsdataverse.org/">wehoop</a></span>',
  '<span>Built with R, <a href="https://quarto.org">Quarto</a>, and a bit of custom HTML and JS</span>',
  "</div>",
  sep = ""
)
```

:::

```{r script}
#| output: asis
cat("<script>\n", paste(readLines("courtside.js", warn = FALSE), collapse = "\n"), "\n</script>\n")
```
